White Dwarf Mergers on Adaptive Meshes\\ I. Methodology and Code Verification

### 2016-01-16 papers/wdmerger_I/paper.tex
Updated paper I

\documentclass[iop,numberedappendix]{../emulateapj}

@@

% allows in-document links; see http://www.astrobetter.com/blog/2014/09/29/latex-hyperref-and-emulateapj/
\usepackage[backref,breaklinks,colorlinks,citecolor=blue]{hyperref} 
\usepackage[all]{hypcap} %Links go to figures;breaks on deluxetables       
\renewcommand*{\backref}[1]{[#1]}

@@

\shorttitle{WD Mergers I. Methodology}
\shortauthors{Katz et al. (2015)}

@@

Another approach to conserving total energy recently taken in the literature is to evolve an 
equation for the total energy $(\rho E_{\text{tot}})$; see \cite{marcello:2012} and \cite{jiang:2013} (see also 
\cite{arepo}, Section 5.3). That is, one can replace the gas energy equation with a total energy equation, 
and then the energy flux includes a term corresponding to the flux of gravitational potential energy. We 
avoid this approach for our problem because there are regions on the computational domain where the total 
energy is dominated by potential energy (especially the low-density regions near the edge of the white dwarfs),
and the gas energy can only be retrieved by first subtracting $-\rho \Phi/2$ from the total energy. Like 
\cite{arepo}, we find that this can result in some serious errors due to numerical discretization, yielding 
unphysical energies or temperatures. We observe also that the implementation of \cite{jiang:2013} 
has terms in the gravitational flux that are not proportional to $\rho$, and so can lead to the 
same troubles that plague the tensor-based formalism for the momentum equation, where small errors 
in the discretization of the gravitational potential can lead to very large changes in the energy of the gas.

@@

Here $R(y)$ is a ramp function that describes the transition between
the two fluids, while $v_{\text{bulk}}$ is the bulk motion of the
fluid in the $y$ direction and $v^\prime$ is the velocity perturbation
that seeds the instability. The problem is established for two
sets of initial conditions (ICs), which we follow
R10 in calling ICs A and B. They differ in
their ramp function ($R_A$ and $R_B$ respectively), as well as the
initial perturbation ($v^\prime_A$ and $v^\prime_B$ respectively), and
the frequency of the perturbation ($n_A = 4$ and $n_B = 2$):
\begin{align}
  R_A &= \begin{cases} 0 & |y - 0.5| > 0.25 \\ 1 & |y - 0.5| < 0.25 \end{cases} \label{eq:kh_ic_a_ramp}\\
  R_B &= \Big\{\left[1 + e^{-2(y-0.25)/\Delta_y}\right]\left[1 + e^{2(y-0.75)/\Delta_y}\right]\Big\}^{-1} \label{eq:kh_ic_b_ramp}\\
  v^\prime_A &= w_0\, \text{sin}\left(n_A\, \pi\, x\right) \left\{e^{-(y-0.25)^2 / 2\sigma^2} + e^{-(y-0.75)^2/2\sigma^2}\right\} \label{eq:kh_ic_a}\\
  v^\prime_B &= w_0\, \text{sin}\left(n_B\, \pi\, x\right). \label{eq:kh_ic_b}
\end{align}

\begin{figure*}[ht]
  \centering
  \includegraphics[trim=0.75in 0.45in 0.85in 0.5in, clip, scale = 0.75]{{{plots/kh_t2.0_p3_low_res_collage}}}
  \caption{2D Kelvin-Helmholtz instability test at $t = 2.0$ for the initial 
    conditions given by \autoref{eq:kh_p3_rho} through \autoref{eq:kh_p3_vy}. 
    The meaning of the rows and columns is the same as in \autoref{fig:kh_p2}.
    \label{fig:kh_p3}  {\color{red} You should note this is $R_B$}}
\end{figure*}

Here $w_0 = 0.1$ is the scale of the velocity perturbation, $\sigma =
0.05/\sqrt{2}$ controls the width of the Gaussian for IC A, and
$\Delta y = 0.05$ is the transition distance scale for the smooth ramp of IC
B. The pressure everywhere is set to $p = 2.5$, and we run this with a
gamma-law equation of state set to $\gamma = 5/3$. Plotfiles are
generated every 0.05 seconds, and the problem is run until $t = 2$.

@@

\citet{mcnally:2012} published another Kelvin-Helmholtz problem that 
is well-posed in the sense that it converges with resolution and 
is not subject to uncontrollable numerical instabilities. Though they 
were not explicitly interested in the question of Galilean invariance, 
we visit that issue here to see what can be learned. The initial 
conditions for this problem are:
\begin{align}
  \rho &= \begin{dcases} \rho_1 - \rho_m e^{(y-0.25)/\Delta y} & 0.25 > y \geq 0 \\ 
                         \rho_2 + \rho_m e^{(0.25-y)/\Delta y} & 0.5 > y \geq 0.25 \\
                         \rho_2 + \rho_m e^{(y-0.75)/\Delta y} & 0.75 > y \geq 0.5 \\
                         \rho_1 - \rho_m e^{(0.75-y)/\Delta y} & 1 > y \geq 0.75 \end{dcases} \label{eq:kh_p3_rho}\\
  v_x &= \begin{dcases} v_1 - v_m e^{(y-0.25)/\Delta y} & 0.25 > y \geq 0 \\
                        v_2 + v_m e^{(0.25-y)/\Delta y} & 0.5 > y \geq 0.25 \\
                        v_2 + v_m e^{(y-0.75)/\Delta y} & 0.75 > y \geq 0.5 \\
                        v_1 - v_m e^{(0.75-y)/\Delta y} & 1 > y \geq 0.75 \end{dcases} \label{eq:kh_p3_vx} \\
  v_y &= w_0\, \text{sin}\left(4\pi x\right). \label{eq:kh_p3_vy}
\end{align}
Here $\Delta y = 0.025$, $w_0 = 0.01$, $v_m = (v_1 - v_2) / 2$, $\rho_m = (\rho_1 - \rho_2) / 2$, 
and the other symbols have the same meaning as above (this means the flow direction
is reversed compared to the original paper, so as to achieve consistency with the 
other simulations presented here). We run this problem at all the same resolutions 
and bulk velocities as the previous two problems. The results for the normal resolutions 
at $t = 2.0$ are displayed in \autoref{fig:kh_p3}. We see a similar pattern as for 
the test proposed by R10: as we get to higher flow speeds we need to have higher 
spatial resolution to compensate for the increased numerical diffusion. The 
accuracy is much lower for the highest bulk velocities. This is probably because 
the initial conditions feature a transition that is twice as sharp (compare the $\Delta y$ 
parameters in each case). This means that there are smaller scale numerical 
perturbations on the grid, which can be amplified by numerical noise in the time 
evolution of the system.

@@

\capstartfalse
\input{plots/sources.table}
\capstarttrue

@@

Turning to the conservation properties of the system, we examine as
fairly typical cases the equal mass system in the inertial frame for
energy conservation (\autoref{fig:energy_conservation_equal}), and the
unequal mass system in the rotating frame for angular momentum
conservation (\autoref{fig:angular_momentum_conservation_unequal}).
For the former system Angular momentum is conserved to within 10 percent over the 25
orbits, while energy conservation is about an order of magnitude
better. We note that while this is already a fairly good level of
energy conservation, it is not nearly as good as the results of
\citet{marcello:2012}. This is because we reset the internal energy to
a level corresponding to our temperature floor when it goes negative,
while \citeauthor{marcello:2012} do not reset and instead ignore the
internal energy if it is negative. The resets impose an artificial
floor on our ability to conserve energy, but they only happen in
low-density regions and do not much affect the large-scale dynamics.
Meanwhile, relative angular momentum conservation is not quite as good 
as relative energy conservation.  This is 
linked to the decline (or increase) in the size of the orbit. This
implies that we ought to be careful in concluding that at these
moderate resolutions we can safely evolve systems for many dozens of
orbits; this needs to be verified to ensure that an observed inspiral
and merger is physically (not numerically) motivated.

@@

Finally we consider whether the dynamical behavior of the system converges with resolution. 
In \autoref{fig:unequal_spatial_convergence} we plot the first 10 seconds of the orbit for 
the unequal mass system, at three different resolutions in the inertial frame: our default 
resolution of $256^3$ zones, as well as a single level of refinement with a jump by a factor of 
two (effective resolution $512^3$) or a jump by a factor of four (effective resolution $1024^3$). 
It is clear that at the latter resolution (corresponding to physical resolution of 100 km), 
we have achieved convergent behavior. In the rotating frame (not shown here), the results are 
not quite the same but for these resolutions the results do tend toward convergence. These 
results suggest a minimum resolution of 200 km for getting the dynamics right.

\begin{figure}
  \centering
  \includegraphics[scale=0.45]{plots/spatial_convergence_rot0}
  \caption{Distance between the two white dwarfs in the unequal mass system, for the first 
           10 seconds of their orbit. The distance is scaled by the initial orbital distance. 
           We plot at three different resolutions, corresponding to the number of 
           effective zones per dimension in the refined regions. Note that the $n = 1024$ curve
           falls nearly exactly on top of the $n = 512$ curve.\label{fig:unequal_spatial_convergence}}
\end{figure}

@@

To examine the parallel performance of \castro, we performed both
strong scaling and weak scaling tests on the Blue Waters machine at
the National Center for Supercomputing Applications. For the weak
scaling test, whose results are shown in \autoref{fig:weak_scaling},
we ran a uniform grid binary white dwarf simulation for resolutions of
$128^3$ zones through $2048^3$ zones. The number of processors was
scaled with the number of zones so that each processor had the same
amount of work; the smallest test used 8 processors and the largest
used 32,768 (note that the number of processor cores on a Blue Waters
node is twice the number of floating point units on that node). The
test was run for 10 timesteps, with each timestep including two
Poisson solves and a hydrodynamics update (though for a uniform grid
calculation we generally do not need to perform any multigrid
iterations for the first Poisson solve in a timestep, since the
density distribution has not changed since the end of the last
timestep). We disabled plotfile and checkpoint writing, as well as
calculation of diagnostic information (the latter can contribute to a
significant fraction of the run time at large processor counts if
computed every timestep). We computed the median wall time required
per time step for each simulation, and then normalized this to the
median time per timestep for the smallest simulation. We find
excellent weak scaling through 4,096 processors. At the largest run,
the simulation time required is slightly than 1.5 times the amount
required for the smallest simulation.  This is due entirely to the
increased cost of the multigrid Poisson solve in each timestep and
this cannot be mitigated except by improving communication or
computation efficiency in the multigrid solver. We observe that this
weak scaling behavior with Poisson gravity is a significant
improvement over the results presented in the first \castro\ paper.

@@

The authors thank John Bell for helpful discussions on gravity 
and hydrodynamics issues. We also thank Noel Scudder and Platon Karpov 
for their help with this project, especially related to visualization of the results, 
and Adam Jacobs for his advice on and assistance with 
running on supercomputing resources. Special thanks are given 
to the organizers of the 2015 Caltech Gravitational Wave 
Astrophysics School, and their supporters: the NSF 
under CAREER award PHY-1151197, the Sherman Fairchild 
Foundation, the LIGO Laboratory at Caltech, and Caltech.

@@

@@

\bibliographystyle{../apj}
\bibliography{../refs}

>>>>

\documentclass[twocolumn,numberedappendix]{../aastex6}

@@

@@

\shorttitle{WD Mergers I. Methodology}
\shortauthors{Katz et al. (2016)}

@@

Another approach to conserving total energy recently taken in the literature is to evolve an 
equation for the total energy $(\rho E_{\text{tot}})$; see \cite{jiang:2013} (see also 
\cite{arepo}, Section 5.3). That is, one can replace the gas energy equation with a total energy equation, 
and then the energy flux includes a term corresponding to the flux of gravitational potential energy. We 
avoid this approach for our problem because there are regions on the computational domain where the total 
energy is dominated by potential energy (especially the low-density regions near the edge of the white dwarfs),
and the gas energy can only be retrieved by first subtracting $-\rho \Phi/2$ from the total energy. Like 
\cite{arepo}, we find that this can result in some serious errors due to numerical discretization, yielding 
unphysical energies or temperatures. We observe also that the implementation of \cite{jiang:2013} 
has terms in the gravitational flux that are not proportional to $\rho$, and so can lead to the 
same troubles that plague the tensor-based formalism for the momentum equation, where small errors 
in the discretization of the gravitational potential can lead to very large changes in the energy of the gas.

@@

Here $R(y)$ is a ramp function that describes the transition between
the two fluids, while $v_{\text{bulk}}$ is the bulk motion of the
fluid in the $y$ direction and $v^\prime$ is the velocity perturbation
that seeds the instability. The problem is established for two
sets of initial conditions (ICs), which we follow
R10 in calling ICs A and B. They differ in
their ramp function ($R_A$ and $R_B$ respectively), as well as the
initial perturbation ($v^\prime_A$ and $v^\prime_B$ respectively), and
the frequency of the perturbation ($n_A = 4$ and $n_B = 2$):
\begin{align}
  R_A &= \begin{cases} 0 & |y - 0.5| > 0.25 \\ 1 & |y - 0.5| < 0.25 \end{cases} \label{eq:kh_ic_a_ramp}\\
  R_B &= \Big\{\left[1 + e^{-2(y-0.25)/\Delta y}\right]\left[1 + e^{2(y-0.75)/\Delta y}\right]\Big\}^{-1} \label{eq:kh_ic_b_ramp}
\end{align}
\begin{align}
  v^\prime_A &= w_0\, \text{sin}\left(n_A\, \pi\, x\right) \left\{e^{-(y-0.25)^2 / 2\sigma^2} + e^{-(y-0.75)^2/2\sigma^2}\right\} \label{eq:kh_ic_a}\\
  v^\prime_B &= w_0\, \text{sin}\left(n_B\, \pi\, x\right). \label{eq:kh_ic_b}
\end{align}
Here $w_0 = 0.1$ is the scale of the velocity perturbation, $\sigma =
0.05/\sqrt{2}$ controls the width of the Gaussian for IC A, and
$\Delta y = 0.05$ is the transition distance scale for the smooth ramp of IC
B. The pressure everywhere is set to $p = 2.5$, and we run this with a
gamma-law equation of state set to $\gamma = 5/3$. Plotfiles are
generated every 0.05 seconds, and the problem is run until $t = 2$.

\begin{figure*}[ht]
  \centering
  \includegraphics[trim=0.75in 0.45in 0.85in 0.5in, clip, scale = 0.75]{{{plots/kh_t2.0_p3_low_res_collage}}}
  \caption{2D Kelvin-Helmholtz instability test at $t = 2.0$ for the initial 
    conditions given by \autoref{eq:kh_p3_rho} through \autoref{eq:kh_p3_vy},
    which come from McNally et al. (2012).
    The meaning of the rows and columns is the same as in \autoref{fig:kh_p2}.
    \label{fig:kh_p3}}
\end{figure*}

@@

\citet{mcnally:2012} published another Kelvin-Helmholtz problem that 
is well-posed in the sense that it converges with resolution and 
is not subject to uncontrollable numerical instabilities. Though they 
were not explicitly interested in the question of Galilean invariance, 
we visit that issue here to see what can be learned. The initial 
conditions for this problem are:
\begin{align}
  \rho &= \begin{dcases} \rho_1 - \rho_m e^{(y-0.25)/\Delta y} & 0.25 > y \geq 0 \\ 
                         \rho_2 + \rho_m e^{(0.25-y)/\Delta y} & 0.5 > y \geq 0.25 \\
                         \rho_2 + \rho_m e^{(y-0.75)/\Delta y} & 0.75 > y \geq 0.5 \\
                         \rho_1 - \rho_m e^{(0.75-y)/\Delta y} & 1 > y \geq 0.75 \end{dcases} \label{eq:kh_p3_rho}
\end{align}
\begin{align}
  v_x &= \begin{dcases} v_1 - v_m e^{(y-0.25)/\Delta y} & 0.25 > y \geq 0 \\
                        v_2 + v_m e^{(0.25-y)/\Delta y} & 0.5 > y \geq 0.25 \\
                        v_2 + v_m e^{(y-0.75)/\Delta y} & 0.75 > y \geq 0.5 \\
                        v_1 - v_m e^{(0.75-y)/\Delta y} & 1 > y \geq 0.75 \end{dcases} \label{eq:kh_p3_vx} \\
  v_y &= w_0\, \text{sin}\left(4\pi x\right). \label{eq:kh_p3_vy}
\end{align}
Here $\Delta y = 0.025$, $w_0 = 0.01$, $v_m = (v_1 - v_2) / 2$, $\rho_m = (\rho_1 - \rho_2) / 2$, 
and the other symbols have the same meaning as above (this means the flow direction
is reversed compared to the original paper, so as to achieve consistency with the 
other simulations presented here). We run this problem at all the same resolutions 
and bulk velocities as the previous two problems. The results for the normal resolutions 
at $t = 2.0$ are displayed in \autoref{fig:kh_p3}. We see a similar pattern as for 
the test proposed by R10: as we get to higher flow speeds we need to have higher 
spatial resolution to compensate for the increased numerical diffusion. The 
qualitative accuracy is much lower for the highest bulk velocities for this problem
than for the previous problems. This is because the amplitude of the instability overall
is smaller than for the previous problems, at least by $t = 2.0$, so it is easier
for numerical diffusion at the shearing layer, caused by the high bulk velocities,
to completely wipe out the instability. Like \citet{robertson:2010} found for
their problem, we find for this problem that the convergence properties are not
substantially affected by altering the perturbation frequency -- the results
show the same qualitative pattern even if we halve this frequency.

@@

\input{plots/sources.table}

@@

Turning to the conservation properties of the system, we examine as
fairly typical cases the equal mass system in the inertial frame for
energy conservation (\autoref{fig:energy_conservation_equal}), and the
unequal mass system in the rotating frame for angular momentum
conservation (\autoref{fig:angular_momentum_conservation_unequal}).
For the former system angular momentum is conserved to within 10 percent over the 25
orbits, while energy conservation is about an order of magnitude
better. We note that while this is already a fairly good level of
energy conservation, it is not nearly as good as the results of
\citet{marcello:2012}. This is because we reset the internal energy to
a level corresponding to our temperature floor when it goes negative,
while \citeauthor{marcello:2012} do not reset and instead ignore the
internal energy if it is negative. The resets impose an artificial
floor on our ability to conserve energy, but they only happen in
low-density regions and do not much affect the large-scale dynamics.
Meanwhile, relative angular momentum conservation is not quite as good 
as relative energy conservation.  This is 
linked to the decline (or increase) in the size of the orbit. This
implies that we ought to be careful in concluding that at these
moderate resolutions we can safely evolve systems for many dozens of
orbits; this needs to be verified to ensure that an observed inspiral
and merger is physically (not numerically) motivated.

@@

Finally we consider whether the dynamical behavior of the system converges with resolution. 
In \autoref{fig:unequal_spatial_convergence_inertial} we plot the first full orbit for 
the unequal mass system, at three different resolutions in the inertial frame: our default 
resolution of $256^3$ zones, as well as a single level of refinement with a jump by a factor of 
two (effective resolution $512^3$) or a jump by a factor of four (effective resolution $1024^3$). 
It is clear that at the latter resolution (corresponding to physical resolution of 100 km), 
we have achieved convergent behavior. In the rotating frame, the results also show
convergent behavior but the convergence is not as fast with resolution as in the inertial frame; 
see \autoref{fig:unequal_spatial_convergence_rotating}. At the two higher resolutions the white dwarf
distance is qualitatively similar, and both are qualitatively different from the lower resolution. However,
quantitatively the two higher resolution runs are not as similar to each other as the analogous runs in the
inertial frame. Convergence with resolution is slightly slower in the rotating reference frame
because in the rotating reference frame a stable, unchanging circular orbit requires balance between
two forces with opposite sign (the gravitational and centrifugal forces), and slight perturbations from the
circular orbit are amplified by the effect of the Corolis force. In the inertial frame, these numerical instabilities
vanish, but the cost is that there is no centrifugal force to actively maintain the white dwarf distance,
which is why it is much more likely for the orbit to prematurely decay. In either case, these results suggest
at least a minimum resolution of 200 km for getting the dynamics qualitatively right. To put that into context,
consider that the parameter study of \citet{dan:2014} used 40,000 SPH particles per simulation, or (for an equal mass
binary) 20,000 particles per white dwarf. For, say, a $0.9\ \msolar + 0.9\ \msolar$ white dwarf binary on
a $256^3$ zone simulation grid, there are 20,000 zones that fit within a white dwarf. We do not
intend here to directly compare results between the two simulation methods. We limit ourselves to the
observation that at least for grid-based codes, a parameter study such as the ones performed by
\citet{dan:2012} and \citet{dan:2014} would likely not yield qualitatively convergent results if it were to use the
same effective mass resolution. Instead the number of zones inside each star should at least be doubled.

\begin{figure}
  \centering
  \includegraphics[scale=0.45]{plots/spatial_convergence_rot0}
  \caption{Distance between the two white dwarfs in the unequal mass system, for the first orbit.
           The distance is scaled by the initial orbital distance. 
           We plot at three different resolutions, corresponding to the number of 
           effective zones per dimension in the refined regions.
           \label{fig:unequal_spatial_convergence_inertial}}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[scale=0.45]{plots/spatial_convergence_rot1}
  \caption{Distance between the two white dwarfs in the unequal mass system, for the first orbit.
           The distance is scaled by the initial orbital distance. 
           We plot at three different resolutions, corresponding to the number of 
           effective zones per dimension in the refined regions.
           \label{fig:unequal_spatial_convergence_rotating}}
\end{figure}

@@

To examine the parallel performance of \castro, we performed both
strong scaling and weak scaling tests on the Blue Waters machine at
the National Center for Supercomputing Applications. For the weak
scaling test, whose results are shown in \autoref{fig:weak_scaling},
we ran a uniform grid binary white dwarf simulation for resolutions of
$128^3$ zones through $2048^3$ zones. The number of processors was
scaled with the number of zones so that each processor had the same
amount of work; the smallest test used 8 processors and the largest
used 32,768 (note that the number of processor cores on a Blue Waters
node is twice the number of floating point units on that node). The
test was run for 10 timesteps, with each timestep including two
Poisson solves and a hydrodynamics update (though for a uniform grid
calculation we generally do not need to perform any multigrid
iterations for the first Poisson solve in a timestep, since the
density distribution has not changed since the end of the last
timestep). We disabled plotfile and checkpoint writing, as well as
calculation of diagnostic information (the latter can contribute to a
significant fraction of the run time at large processor counts if
computed every timestep). We computed the median wall time required
per time step for each simulation, and then normalized this to the
median time per timestep for the smallest simulation. We find
excellent weak scaling through 4,096 processors. At the largest run,
the simulation time required is slightly less than 1.5 times the amount
required for the smallest simulation.  This is due entirely to the
increased cost of the multigrid Poisson solve in each timestep and
this cannot be mitigated except by improving communication or
computation efficiency in the multigrid solver. We observe that this
weak scaling behavior with Poisson gravity is a significant
improvement over the results presented in the first \castro\ paper.

@@

The authors thank John Bell, Volker Springel, and Dominic Marcello
for helpful discussions on gravity 
and hydrodynamics issues. We also thank Noel Scudder and Platon Karpov 
for their help with this project, especially related to visualization of the results, 
and Adam Jacobs for his advice on and assistance with 
running on supercomputing resources. Special thanks are given 
to the organizers of the 2015 Caltech Gravitational Wave 
Astrophysics School, and their supporters: the NSF 
under CAREER award PHY-1151197, the Sherman Fairchild 
Foundation, the LIGO Laboratory at Caltech, and Caltech.

@@

\software{BoxLib (\url{https://github.com/BoxLib-Codes/BoxLib}),
          CASTRO (\citet{castro}, \url{https://github.com/BoxLib-Codes/Castro}),
          wdmerger (\url{https://github.com/BoxLib-Codes/wdmerger}),
          GCC (\url{https://gcc.gnu.org/}),
          python (\url{https://www.python.org/}),
          matplotlib (\citet{matplotlib}, \url{http://matplotlib.org/}),
          yt (\citet{yt}, \url{http://yt-project.org/})}

@@

\bibliographystyle{../aasjournal}
\bibliography{../refs}
